Co-methylation

### 2015-05-02 Rmarkdown/comethylation.Rmd
Address Terry’s suggestions for co-methlation chapter.

The second strategy uses all pairs of CpGs on the same chromosome and strand, which are then stratified by $IPD$ and any secondary variables, such as genomic context. In practice, the second strategy only uses pairs from a set of $IPD$s, e.g., $IPD = 2 - 2000$, otherwise the number of pairs becomes unwieldy. This is more like a traditional autocorrelation analysis of $\bm{\beta} = (\beta_{1}, \ldots, \beta_{N_{loci}})$, albeit of a vector of unequally spaced observations and with care taken to only use pairs on the same chromosome and strand. I refer to this as the $NIL \geq 0$ strategy. Using the $NIL \geq 0$ strategy, two pairs of CpGs with an identical $IPD$ may have a very different number of intervening CpGs.

@@

One final aspect of these genome-wide plots bears mentioning. There are hints of an approximately $150$ to $200$ bp periodicity in these plots. This is a similar scale to the previously reported periodicities in correlations of $\beta$-values (see Chapter \ref{chap:co-methylation_review}). While this periodicity can readily be detected by eye, it is somewhat more difficult to verify using traditional methods based on Fourier analysis of the signal. The chief reason is that the data are very noisy beyond the second or third putative cycle due to the limited data at these larger $IPD$s.

@@

Figures \ref{fig:EPISCOPE_spearman_beta_cors_min_cov_10_strand_collapsed_CI}, \ref{fig:Lister_spearman_beta_cors_min_cov_10_strand_collapsed_CI}, \ref{fig:Seisenberger_spearman_beta_cors_min_cov_10_strand_collapsed_CI} and \ref{fig:Ziller_spearman_beta_cors_min_cov_10_strand_collapsed_CI} are plots of Spearman correlations of $\beta$-values for pairs of CpGs stratified by whether the pair is inside or outside of a CpG island[^spanning]. Again, the substantial noise in these plots is due to a lack of $NIL = 0$ pairs with large $IPD$s, particularly inside CpG islands.

@@

The odds ratio is not perfece, however. For one, the possible values of $\psi$ are highly skewed, with $0 < \psi < 1$ and $1 < \psi < \infty$ corresponding to the two distinct dependence scenarios. Rather than estimating $\psi$ directly, we will instead focus our attention on the _log odds ratio_, $\theta = \log{\psi}$. The log odds ratio is symmetric about 0, which makes statistical inference somewhat simpler. I will routinely switch between discussing the estimation of $\psi$ and $\theta$, noting that we can always convert one to the other by appropriate exponentiation or taking of logarithms.

@@

Another alternative to $\widehat{\theta_{U}}$ is the _conditional_ maximum likelihood estimator, $\widehat{\theta_{C}}$. This is computed by first conditioning on ${n_{1 +}, n_{+ 1}}$. Some algebra then leads to the conditional distribution of $n_{11}$, $f(n_{11}; n_{1 +}, n_{+, 1, \psi})$, which is a hypergeometric distribution \citep{Fisher:1935ug}. The maximum likelihood estimator of $\psi$, $\widehat{\psi_{C}}$ is solved using iterative methods. The conditional estimator, $\widehat{\theta_{C}} = \log{\widehat{\psi_{C}}}$, works better than the unconditional estimator, $\widehat{\theta_{U}}$, when the sample size, $n$, is small \citep{Agresti:2007tf}.

@@

We can again compute an unconditional maximum likelihood estimator, $\widehat{\psi_{U}}$, a '0.5-adjusted' unconditional maximum likelihood estimator, $\widehat{\psi_{0.5}}$, and a conditional maximum likelihood estimator, $\widehat{\psi_{C}}$, of the log odds ratio $\theta = \log{\psi}$ \citep[see][]{Breslow:1981ta}.

Another estimator of the common odds ratio, $\psi$, is the Mantel-Haenszel estimator \citep{MANTEL:1959uf}, $\widehat{\psi_{MH}} = \frac{\sum_{k} (n_{11k} n_{22k} / n_{+ + k})}{n_{12k} n_{21k} / n_{+ + k}}$. \citet{Robins:1986vh} derived a simple robust estimator of the variance for $\widehat{\theta}_{MH} = \log{\widehat{\psi_{MH}}}$, $\hat{\sigma}^{2}(\widehat{\theta}_{MH})$. This variance estimator did not appear until some 25 years after the initial publication of the Mantel-Haenszel estimator[^mh_variance]. This delay was in part due to the two asymptotic forms of $2 \times 2 \times K$ contingency tables.

@@

\citet{Breslow:1981ta} showed that under $A2$ neither the unconditional maximum likelihood estimator, $\widehat{\psi_{U}}$, nor the '0.5-adjusted' unconditional maximum likelihood estimator, $\widehat{\psi_{0.5}}$, converge to the true odds ratio. In contrast, both the Mantel-Haenszel estimator, $\widehat{\psi_{MH}}$, and the conditional maximum likelihood estimator, $\widehat{\psi_{C}}$, do convert to $\psi$, with $\widehat{\psi_{MH}}$ having the advantage of a simple closer form expression \citep{Breslow:1996uc}.

@@

We analyse the results of a simulation study to explore the effects of sparsity and heterogeneity of odds ratios on estimates of odds ratios in $2 \times 2 \times K$ contingency tables. The simulation requires that we specify several parameters. In order to relate these back to DNA methylation, I describe each parameter using the statistical framework proposed in Chapter \ref{chap:wgbs_statistical_framework}:

- the number of tables, $K$. The number of 2-tuples.
- the marginal probabilities, $\pi_{2 +} = Pr(Z_{i})$ and $\pi_{+ 2} = Pr(Z_{i'})$. Some simulations use a common set of marginal probabilities across all tables but others simulate the marginal probabilities from a specified distribution.
- the odds ratio of each table, $\psi_{k}$. Most simulations use a common odds ratio across the $K$ tables, $\psi_{k} = \psi$.
- the sample size of each table, $n_{+ + k} = d_{(i, i')}$. The number of reads containing both the $i^{th}$ and ${i'}^{th}$ methylation locus (sequencing depth). Most simulations use a truncated negative binomial distribution to model the sequencing depth. The truncated negative binomial distribution is parameterised by the mean ($\mu$), the dispersion parameter (size) and the truncation level (trunc). Unless otherwise noted, simulations use $\mu = 30$, $\text{size} = 10$ and $\text{trunc} = 9$, corresponding to an average sequencing depth of $30 \times$ but where only 2-tuples with at least $10 \times$ sequencing depth are used in downstream analyses[^nb_sim].

@@

1. The $K$ 0.5-adjusted unconditional odds ratio estimates, $\widehat{\psi_{0.5}}$, from each $2 \times 2$ table.
2. The unconditional odds ratio estimate from the $2 \times 2$ table formed by collapsing over $k$, $\widehat{\psi_{U}}$.
3. The Mantel-Haenszel odds ratio estimate from the $2 \times 2 \times K$ table, $\widehat{\psi_{MH}}$.

The results of each simulation are summarised by a plot of $\widehat{\theta_{0.5}} = \log(\widehat{\psi_{0.5}})$ (histogram), the point estimate (solid line) and $95 \%$ confidence interval (dashed lines) of $\widehat{\theta_{U}} = \log(\widehat{\psi_{U}})$ (coloured orange), and the point estimate (solid line) and $95 \%$ confidence interval (dashed lines) of $\widehat{\theta}_{MH} = \log(\widehat{\psi_{MH}})$ (coloured blue). The density of the true value of the log odds ratio across the $2 \times 2 \times K$ table $\theta$, is shown by a dashed black line[^density].

[^density]: In many simulations $\psi$ is constant across the $K$ tables and therefore the density is degenerate.

@@

To begin, we explore the effect of the marginal probabilities, $\pi_{2 +}, \pi_{+ 2}$, when $\theta = 0$ ($\psi = 1$) identically across the $K$ tables (corresponding to independence of $X$ and $Y$) for $K = 100$ (Simulation 1, Figure \ref{fig:Simulation_1}), and $K = 1000$ (Simulation 2a, Figure \ref{fig:Simulation_2}). Under this simulation, we expect both $\widehat{\theta_{U}}$ and $\widehat{\theta}_{MH}$ to converge to the true value, $\theta = 0$, as $K \to \infty$. This is indeed what we observe, except when the marginal probabilities are uniformly near the boundaries ($\pi_{2 +} = \pi_{+ 2} = 0.01$ and $\pi_{2 +} = \pi_{+ 2} = 0.99$) and $K$ is smaller.

By contrast, $\widehat{\theta_{0.5}}$ are highly biased estimates of $\theta$, even when $K = 1000$. The histograms of $\widehat{\theta_{0.5}}$ are not centred around the true value, $\theta = 0$, unless the marginal probabilities are well away from the boundaries ($0.25 < \pi_{2 +} = \pi_{+ 2} < 0.75$). The bias can be eliminated only by having ultra-deep sequencing coverage ($mean(d) = 10000 \times$), although the variation of $\widehat{\theta_{0.5}}$ across the $K = 1000$ tables remains non-negligible (Simulation 2b Figure \ref{fig:Simulation_2d}).

\begin{figure}[!htb]
\centering
\includegraphics[width=\textwidth]{../figures/Simulation_1.pdf}
\caption[Simulation 1]{Simulation 1. $\widehat{\theta_{0.5}} = \log(\widehat{\psi_{0.5}})$ (histogram), the point estimate (solid line) and $95 \%$ confidence interval (dashed lines) of $\widehat{\theta_{U}} = \log(\widehat{\psi_{U}})$ (coloured orange), and the point estimate (solid line) and $95 \%$ confidence interval (dashed lines) of $\widehat{\theta}_{MH} = \log(\widehat{\psi_{MH}})$ (coloured blue). The density of the true value of the log odds ratio across the $2 \times 2 \times K$ table $\theta$, is shown by a dashed black line. See main text for description of other parameters.}
\label{fig:Simulation_1}
\end{figure}

\begin{figure}[!htb]
\centering
\includegraphics[width=\textwidth]{../figures/Simulation_2.pdf}
\caption[Simulation 2a]{Simulation 2a. $\widehat{\theta_{0.5}} = \log(\widehat{\psi_{0.5}})$ (histogram), the point estimate (solid line) and $95 \%$ confidence interval (dashed lines) of $\widehat{\theta_{U}} = \log(\widehat{\psi_{U}})$ (coloured orange), and the point estimate (solid line) and $95 \%$ confidence interval (dashed lines) of $\widehat{\theta}_{MH} = \log(\widehat{\psi_{MH}})$ (coloured blue). The density of the true value of the log odds ratio across the $2 \times 2 \times K$ table $\theta$, is shown by a dashed black line. See main text for description of other parameters.}
\label{fig:Simulation_2}
\end{figure}

\begin{figure}[!htb]
\centering
\includegraphics[width=\textwidth]{../figures/Simulation_2d.pdf}
\caption[Simulation 2b]{Simulation 2b. $\widehat{\theta_{0.5}} = \log(\widehat{\psi_{0.5}})$ (histogram), the point estimate (solid line) and $95 \%$ confidence interval (dashed lines) of $\widehat{\theta_{U}} = \log(\widehat{\psi_{U}})$ (coloured orange), and the point estimate (solid line) and $95 \%$ confidence interval (dashed lines) of $\widehat{\theta}_{MH} = \log(\widehat{\psi_{MH}})$ (coloured blue). The density of the true value of the log odds ratio across the $2 \times 2 \times K$ table $\theta$, is shown by a dashed black line. See main text for description of other parameters.}
\label{fig:Simulation_2d}
\end{figure}

Next we explore what happens when we allow the marginal probabilities, $\pi_{2 +}, \pi_{+ 2}$, to vary across the $K$ tables when $\theta$ is a fixed value identically across the $K$ tables for $K = 100$ ($\theta$ not necessarily zero). To do so, we simulate $\pi_{2 +}, \pi_{+ 2} \eqd Uniform(0, 1)$ for each $2 \times 2$ table. As is expected from the theory, Figure \ref{fig:Simulation_3} shows that the Mantel-Haenszel estimator does an excellent job of estimating the true value of $\theta$. By contrast, $\widehat{\theta_{U}}$, is well away from the true value and the distribution of $\widehat{\theta_{0.5}}$ is skewed and widely dispersed.

\begin{figure}[!htb]
\centering
\includegraphics[width=\textwidth]{../figures/Simulation_3.pdf}
\caption[Simulation 3]{Simulation 3. $\widehat{\theta_{0.5}} = \log(\widehat{\psi_{0.5}})$ (histogram), the point estimate (solid line) and $95 \%$ confidence interval (dashed lines) of $\widehat{\theta_{U}} = \log(\widehat{\psi_{U}})$ (coloured orange), and the point estimate (solid line) and $95 \%$ confidence interval (dashed lines) of $\widehat{\theta}_{MH} = \log(\widehat{\psi_{MH}})$ (coloured blue). The density of the true value of the log odds ratio across the $2 \times 2 \times K$ table $\theta$, is shown by a dashed black line. See main text for description of other parameters.}
\label{fig:Simulation_3}
\end{figure}

In a sense, the $\pi_{2 +}, \pi_{+ 2} \eqd Uniform(0, 1)$ assumption is too easy. As we have seen in Chapter \ref{chap:wgbs_statistical_framework}, the marginal probabilities of DNA methylation are mostly near the boundaries, not uniformly distributed on the unit interval. Furthermore, each methylation locus in a 2-tuple will likely have a similar marginal probability of methylation. To simulate this more realistic scenario, we simulate $\pi_{2 +} = \pi_{+ 2} \eqd Beta(0.3, 0.2)$, which has a similar shape to the genome-wide distribution of $\beta$-values observed in most mammalian genomes (Figure \ref{fig:beta_marginals}). The results of this simulation (Simulation 4, Figure \ref{fig:Simulation_4}) are basically an exaggerated version of Simulation 3. The Mantel-Haenszel estimator, $\widehat{\theta}_{MH}$, remains an excellent estimator; the unconditional log odds ratio estimator of the collapsed table does an awful job, regularly returning $\widehat{\theta_{U}} > 0$ when in fact $\theta < 0$; and the distribution of $\widehat{\theta_{0.5}}$ is even more widely dispersed and typically not centred around $\theta$.

@@

\begin{figure}[!htb]
\centering
\includegraphics[width=\textwidth]{../figures/Simulation_4.pdf}
\caption[Simulation 4]{Simulation 4. $\widehat{\theta_{0.5}} = \log(\widehat{\psi_{0.5}})$ (histogram), the point estimate (solid line) and $95 \%$ confidence interval (dashed lines) of $\widehat{\theta_{U}} = \log(\widehat{\psi_{U}})$ (coloured orange), and the point estimate (solid line) and $95 \%$ confidence interval (dashed lines) of $\widehat{\theta}_{MH} = \log(\widehat{\psi_{MH}})$ (coloured blue). The density of the true value of the log odds ratio across the $2 \times 2 \times K$ table $\theta$, is shown by a dashed black line. See main text for description of other parameters.}
\label{fig:Simulation_4}
\end{figure}

We now explore what happens when the odds ratio is heterogeneous across the $K$ tables. Although results are shown for $\widehat{\theta_{0.5}}$ and $\widehat{\theta_{U}}$, we do not discuss these estimators further since we have seen that these perform poorly in even the simplest simulation scenarios.

In these remaining simulations, all parameters are as for Simulation 4 except that the true log odds ratio of each table, $\theta_{k}$, comes from the following distributions:

- Simulation 5 and Simulation 6: $\theta_{k} \eqd Gaussian(\phi_{0}, 1)$, where $\theta_{0} = -4, \ldots, 4$ is identical across the $K$ tables.
- Simulation 7:$\theta_{k} \eqd Skewed\text{-}Gaussian(\theta_{0}, 1, -10)$, where $\theta_{0} = -4, \ldots, 4$ is identical across the $K$ tables.
- Simulation 8: $\theta_{k} \eqd 0.5 \times Gaussian(- \theta_{0}, 1) + 0.5 \times Gaussian(\theta_{0}, 1)$.

Simulations 5, 6 and 7 (Figures \ref{fig:Simulation_5}, \ref{fig:Simulation_6} and \ref{fig:Simulation_7}) demonstrate that the Mantel-Haenszel estimator, $\widehat{\theta}_{MH}$, is a reasonable estimator of the centre of the distribution of log odds ratios. It performs better in Simulation 5 and 6, where the distribution of $\theta$ is symmetric, as opposed to Simulation 7, where the distribution of $\theta$ is skewed. Unsurprisingly, $\widehat{\theta}_{MH}$ does a better job as $K \to \infty$. It should be emphasised, however, that the asymptotic variance of the estimator, $\hat{\sigma}(\widehat{\theta}_{MH})^{2}$ is not an estimate of the variance of the true odds ratios, $var(\theta)$. We can see that the asymptotic $95 \%$ confidence interval of $\widehat{\theta}_{MH}$ only covers a small amount of the variation in the $\theta$.

\begin{figure}[!htb]
\centering
\includegraphics[width=\textwidth]{../figures/Simulation_5.pdf}
\caption[Simulation 5]{Simulation 5. $\widehat{\theta_{0.5}} = \log(\widehat{\psi_{0.5}})$ (histogram), the point estimate (solid line) and $95 \%$ confidence interval (dashed lines) of $\widehat{\theta_{U}} = \log(\widehat{\psi_{U}})$ (coloured orange), and the point estimate (solid line) and $95 \%$ confidence interval (dashed lines) of $\widehat{\theta}_{MH} = \log(\widehat{\psi_{MH}})$ (coloured blue). The density of the true value of the log odds ratio across the $2 \times 2 \times K$ table $\theta$, is shown by a dashed black line. See main text for description of other parameters.}
\label{fig:Simulation_5}
\end{figure}

>>>>

The second strategy uses all pairs of CpGs on the same chromosome and strand, which are then stratified by $IPD$ and any secondary variables, such as genomic context. In practice, the second strategy only uses pairs from a set of $IPD$s, e.g., $IPD = 2 - 2000$, otherwise the number of pairs becomes unwieldy. This is what is called an 'autocorrelation' analysis of $\bm{\beta} = (\beta_{1}, \ldots, \beta_{N_{loci}})$ by \citet{Cokus:2008fc}, although it is somewhat different to the traditional notion of an autocorrelation analysis. I refer to this as the $NIL \geq 0$ strategy. Using the $NIL \geq 0$ strategy, two pairs of CpGs with an identical $IPD$ may have a very different number of intervening CpGs.

@@

One final aspect of these genome-wide plots bears mentioning. There are hints of an approximately $150$ to $200$ bp periodicity in these plots. This is a similar scale to the previously reported periodicities in correlations of $\beta$-values (see Chapter \ref{chap:co-methylation_review}). While this periodicity can readily be detected by eye, it is somewhat more difficult to verify using classical methods from spectral analysis. The chief reason is that the data are very noisy beyond the second or third putative cycle due to the limited data at these larger $IPD$s.

@@

Figures \ref{fig:EPISCOPE_spearman_beta_cors_min_cov_10_strand_collapsed_CGI}, \ref{fig:Lister_spearman_beta_cors_min_cov_10_strand_collapsed_CGI}, \ref{fig:Seisenberger_spearman_beta_cors_min_cov_10_strand_collapsed_CGI} and \ref{fig:Ziller_spearman_beta_cors_min_cov_10_strand_collapsed_CGI} are plots of Spearman correlations of $\beta$-values for pairs of CpGs stratified by whether the pair is inside or outside of a CpG island[^spanning]. Again, the substantial noise in these plots is due to a lack of $NIL = 0$ pairs with large $IPD$s, particularly inside CpG islands.

@@

The odds ratio is a simple summary of the dependence in a $2 \times 2$ table. An attractive feature is that it is independent of the marginal probabilities (__TODO: cite__). The odds ratio is not perfect, however. For one, the possible values of $\psi$ are highly skewed, with $0 < \psi < 1$ and $1 < \psi < \infty$ corresponding to the two distinct dependence scenarios. Rather than estimating $\psi$ directly, we will instead focus our attention on the _log odds ratio_, $\theta = \log{\psi}$. The log odds ratio is symmetric about 0, which makes statistical inference somewhat simpler. I will routinely switch between discussing the estimation of $\psi$ and $\theta$, noting here that we can always convert one to the other by appropriate exponentiation or taking of logarithms.

@@

Another alternative to $\widehat{\theta_{U}}$ is the _conditional_ maximum likelihood estimator, $\widehat{\theta_{C}}$. This is computed by first conditioning on ${n_{1 +}, n_{+ 1}}$. Some algebra then leads to the conditional distribution of $n_{11}$, $f(n_{11}; n_{1 +}, n_{+, 1, \psi})$, which is a hypergeometric distribution \citep{Fisher:1935ug}. The maximum likelihood estimator of $\psi$, $\widehat{\psi}_{C}$ is solved using iterative methods. The conditional estimator, $\widehat{\theta_{C}} = \log{\widehat{\psi}_{C}}$, works better than the unconditional estimator, $\widehat{\theta_{U}}$, when the sample size, $n$, is small \citep{Agresti:2007tf}.

@@

We can again compute an unconditional maximum likelihood estimator, $\widehat{\psi}_{U}$, a '0.5-adjusted' unconditional maximum likelihood estimator, $\widehat{\psi}_{0.5}$, and a conditional maximum likelihood estimator, $\widehat{\psi}_{C}$, of the log odds ratio $\theta = \log{\psi}$ \citep[see][]{Breslow:1981ta}.

Another estimator of the common odds ratio, $\psi$, is the Mantel-Haenszel estimator \citep{MANTEL:1959uf}, $\widehat{\psi}_{MH} = \frac{\sum_{k} (n_{11k} n_{22k} / n_{+ + k})}{n_{12k} n_{21k} / n_{+ + k}}$. \citet{Robins:1986vh} derived a simple robust estimator of the variance for $\widehat{\theta}_{MH} = \log{\widehat{\psi}_{MH}}$, $\hat{\sigma}^{2}(\widehat{\theta}_{MH})$. This variance estimator did not appear until some 25 years after the initial publication of the Mantel-Haenszel estimator[^mh_variance]. This delay was in part due to the two asymptotic forms of $2 \times 2 \times K$ contingency tables.

@@

\citet{Breslow:1981ta} showed that under $A2$ neither the unconditional maximum likelihood estimator, $\widehat{\psi}_{U}$, nor the '0.5-adjusted' unconditional maximum likelihood estimator, $\widehat{\psi}_{0.5}$, converge to the true odds ratio. In contrast, both the Mantel-Haenszel estimator, $\widehat{\psi}_{MH}$, and the conditional maximum likelihood estimator, $\widehat{\psi}_{C}$, do converge to $\psi$, with $\widehat{\psi}_{MH}$ having the advantage of a simple closer form expression \citep{Breslow:1996uc}.

@@

We carry out a simulation study to explore the effects of sparsity and heterogeneity of odds ratios on estimates of odds ratios in $2 \times 2 \times K$ contingency tables. The simulation requires that we specify several parameters. In order to relate these back to DNA methylation, I describe each parameter using the statistical framework proposed in Chapter \ref{chap:wgbs_statistical_framework}:

- the number of tables, $K$. The number of CpG pairs (2-tuples).
- the marginal probabilities, $\pi_{2 +} = Pr(Z_{i})$ and $\pi_{+ 2} = Pr(Z_{i'})$. Some simulations use a common set of marginal probabilities across all tables but others simulate the marginal probabilities from a specified distribution.
- the odds ratio of each table, $\psi_{k}$. Most simulations use a common odds ratio across the $K$ tables, $\psi_{k} = \psi$.
- the sample size of each table, $n_{+ + k} = d_{(i, i')}$. The number of reads containing both the $i^{th}$ and ${i'}^{th}$ methylation locus (sequencing depth). Most simulations use a truncated negative binomial distribution to model the sequencing depth. The truncated negative binomial distribution is parameterised by the mean ($\mu$), the dispersion parameter (size) and the truncation level (trunc). Unless otherwise noted, simulations use $\mu = 30$, $\text{size} = 10$ and $\text{trunc} = 9$, corresponding to an average sequencing depth of $30 \times$ but where only pairs with at least $10 \times$ sequencing depth are used in downstream analyses[^nb_sim].

@@

1. The $K$ 0.5-adjusted unconditional odds ratio estimates, $\widehat{\psi}_{0.5}$, from each $2 \times 2$ table.
2. The unconditional odds ratio estimate from the $2 \times 2$ table formed by collapsing over $k$, $\widehat{\psi}_{U}$.
3. The Mantel-Haenszel odds ratio estimate from the $2 \times 2 \times K$ table, $\widehat{\psi}_{MH}$.

The results of each simulation are summarised by a plot of $\widehat{\theta}_{0.5} = \log(\widehat{\psi}_{0.5})$ (histogram), the point estimate (solid line) and $95 \%$ confidence interval (dashed lines) of $\widehat{\theta}_{U} = \log(\widehat{\psi}_{U})$ (coloured orange), and the point estimate (solid line) and $95 \%$ confidence interval (dashed lines) of $\widehat{\theta}_{MH} = \log(\widehat{\psi}_{MH})$ (coloured blue). The density of the true value of the log odds ratio across the $2 \times 2 \times K$ table $\theta$, is shown by a dashed black line[^density].

[^density]: In many simulations $\theta$ is constant across the $K$ tables and therefore the density is degenerate.

@@

To begin, we explore the effect of the marginal probabilities, $\pi_{2 +}, \pi_{+ 2}$, when $\theta = 0$ ($\psi = 1$) identically across the $K$ tables (corresponding to independence of $X$ and $Y$) for $K = 100$ (Simulation 1a, Figure \ref{fig:Simulation_1a}), and $K = 1000$ (Simulation 1b, Figure \ref{fig:Simulation_1b}). Under this simulation, we expect both $\widehat{\theta_{U}}$ and $\widehat{\theta}_{MH}$ to converge to the true value, $\theta = 0$, as $K \to \infty$. This is indeed what we observe, except when the marginal probabilities are uniformly near the boundaries ($\pi_{2 +} = \pi_{+ 2} = 0.01$ and $\pi_{2 +} = \pi_{+ 2} = 0.99$) and $K$ is smaller. By contrast, the estimates of $\theta$ from the individual $2 \times 2$ tables, $\widehat{\theta_{0.5}}$, are highly biased, even when $K = 1000$ in Simulate 1b (Figure \ref{fig:Simulation_1b}). In both Simulation 1a and 1b (Figures \ref{fig:Simulation_1a} and \ref{fig:Simulation_1b}) the histograms of $\widehat{\theta}_{0.5}$ are not centred around the true value, $\theta = 0$, unless the marginal probabilities are well away from the boundaries ($0.25 < \pi_{2 +} = \pi_{+ 2} < 0.75$).

\begin{figure}[!htb]
\centering
\includegraphics[width=\textwidth]{../figures/Simulation_1.pdf}
\caption[Simulation 1a]{Results of simulation 1a ($K = 100$). Estimating the log odds ratio, $\theta = 0$ (shown by the dashed line), from a $2 \times 2 \times 100$ contingency table when the marginal probabilities, $\pi_{2+}$ and $\pi_{+2}$, are equal. The results of three estimators of $\theta$ are shown: $\widehat{\theta}_{0.5}$ (histogram); the point estimate (solid line) and $95 \%$ confidence interval (dashed lines) of $\widehat{\theta}_{U}$ (coloured orange); and the point estimate (solid line) and $95 \%$ confidence interval (dashed lines) of $\widehat{\theta}_{MH}$ (coloured blue). The average sample size (sequencing-coverage) of each $2 \times 2$ table is $30$.}
\label{fig:Simulation_1a}
\end{figure}

\begin{figure}[!htb]
\centering
\includegraphics[width=\textwidth]{../figures/Simulation_2.pdf}
\caption[Simulation 1b]{Results of simulation 1b ($K = 1000$). Estimating the log odds ratio, $\theta = 0$ (shown by the dashed line), from a $2 \times 2 \times 1000$ contingency table when the marginal probabilities, $\pi_{2+}$ and $\pi_{+2}$, are equal. The results of three estimators of $\theta$ are shown: $\widehat{\theta}_{0.5}$ (histogram); the point estimate (solid line) and $95 \%$ confidence interval (dashed lines) of $\widehat{\theta}_{U}$ (coloured orange); and the point estimate (solid line) and $95 \%$ confidence interval (dashed lines) of $\widehat{\theta}_{MH}$ (coloured blue). The average sample size (sequencing-coverage) of each $2 \times 2$ table is $30$.}
\label{fig:Simulation_1b}
\end{figure}

The bias can be eliminated only by having ultra-deep sequencing coverage, such as in Simulation 1c where $mean(d) = 10000 \times$ (Figure \ref{fig:Simulation_1c}). Even then, however, the variation of $\widehat{\theta}_{0.5}$ across the $K = 1000$ tables remains non-negligible when the marginal probabilities, $\pi_{2+}$ and $\pi_{+2}$ are near the boundaries.

\begin{figure}[!htb]
\centering
\includegraphics[width=\textwidth]{../figures/Simulation_2d.pdf}
\caption[Simulation 1c]{Results of simulation 1c ($K = 1000$). Estimating the log odds ratio, $\theta = 0$ (shown by the dashed line), from a $2 \times 2 \times 1000$ contingency table when the marginal probabilities, $\pi_{2+}$ and $\pi_{+2}$, are equal. The results of three estimators of $\theta$ are shown: $\widehat{\theta}_{0.5}$ (histogram); the point estimate (solid line) and $95 \%$ confidence interval (dashed lines) of $\widehat{\theta}_{U}$ (coloured orange); and the point estimate (solid line) and $95 \%$ confidence interval (dashed lines) of $\widehat{\theta}_{MH}$ (coloured blue). The average sample size (sequencing-coverage) of each $2 \times 2$ table is $10,000$.}
\label{fig:Simulation_1c}
\end{figure}

Next, we explore what happens when we allow the marginal probabilities, $\pi_{2 +}$ and $\pi_{+ 2}$, to vary across the $K = 100$ tables for a fixed (possibly non-zero) value of $\theta$. To do so, we simulate $\pi_{2 +}, \pi_{+ 2} \eqd Uniform(0, 1)$ for each $2 \times 2$ table. As is expected from the theory, the results of Simulation 2 show that the Mantel-Haenszel estimator does an excellent job of estimating the true value of $\theta$ (Figure \ref{fig:Simulation_3}). By contrast, the estimate formed by collapsing over the $K$ tables, $\widehat{\theta}_{U}$, is well away from the true value and the distribution of $\widehat{\theta}_{0.5}$ is skewed and widely dispersed.

\begin{figure}[!htb]
\centering
\includegraphics[width=\textwidth]{../figures/Simulation_3.pdf}
\caption[Simulation 2]{Results of simulation 2 ($K = 100$). Estimating the log odds ratio, $\theta$ (shown by the dashed line), from a $2 \times 2 \times 100$ contingency table where the marginal probabilities, $\pi_{2+}$ and $\pi_{+2}$, are distributed as $Uniform(0, 1)$ random variables. The results of three estimators of $\theta$ are shown: $\widehat{\theta}_{0.5}$ (histogram); the point estimate (solid line) and $95 \%$ confidence interval (dashed lines) of $\widehat{\theta}_{U}$ (coloured orange); and the point estimate (solid line) and $95 \%$ confidence interval (dashed lines) of $\widehat{\theta}_{MH}$ (coloured blue). The average sample size (sequencing-coverage) of each $2 \times 2$ table is $30$.}
\label{fig:Simulation_2}
\end{figure}

In a sense, the $\pi_{2 +}, \pi_{+ 2} \eqd Uniform(0, 1)$ assumption in Simulation 2 is too easy. As we have seen in Chapter \ref{chap:wgbs_statistical_framework}, the marginal probabilities of DNA methylation are not uniformly distributed but, rather, are bimodal and mostly near the boundaries. Furthermore, each methylation locus in a pair will likely have a similar marginal probability of methylation. To simulate this more relevant scenario, we simulate $\pi_{2 +} = \pi_{+ 2} \eqd Beta(0.3, 0.2)$, which has a similar shape to the genome-wide distribution of $\beta$-values observed in most mammalian genomes (Figure \ref{fig:beta_marginals}).

@@

The results of this simulation (Simulation 3, Figure \ref{fig:Simulation_3}) are basically an exaggerated version of Simulation 2. The Mantel-Haenszel estimator, $\widehat{\theta}_{MH}$, remains an excellent estimator; the unconditional log odds ratio estimator of the collapsed table does an awful job, regularly returning $\widehat{\theta}_{U} > 0$ when in fact $\theta < 0$; and the distribution of $\widehat{\theta}_{0.5}$ is even more widely dispersed and typically not centred around $\theta$.

\begin{figure}[!htb]
\centering
\includegraphics[width=\textwidth]{../figures/Simulation_3.pdf}
\caption[Simulation 3]{Results of simulation 3 ($K = 100$). Estimating the log odds ratio, $\theta$ (shown by the dashed line), from a $2 \times 2 \times 100$ contingency table where the marginal probabilities, $\pi_{2+}$ and $\pi_{+2}$, are distributed as $Beta(0.3, 0.2)$ random variables. The results of three estimators of $\theta$ are shown: $\widehat{\theta}_{0.5}$ (histogram); the point estimate (solid line) and $95 \%$ confidence interval (dashed lines) of $\widehat{\theta}_{U}$ (coloured orange); and the point estimate (solid line) and $95 \%$ confidence interval (dashed lines) of $\widehat{\theta}_{MH}$ (coloured blue). The average sample size (sequencing-coverage) of each $2 \times 2$ table is $30$.}
\label{fig:Simulation_3}
\end{figure}

We now turn our attention to what happens when the odds ratio is heterogeneous across the $K$ tables. Although results are shown for $\widehat{\theta}_{0.5}$ and $\widehat{\theta}_{U}$, we do not discuss these estimators further since we have seen that these perform poorly in even the simplest simulation scenarios.

In these remaining simulations, all parameters are as for Simulation 3 except that the true log odds ratio of each table, $\theta_{k}$, comes from the following distributions:

- Simulation 4a and Simulation 4b: $\theta_{k} \eqd Gaussian(\phi_{0}, 1)$, where $\theta_{0} = -4, \ldots, 4$ is identical across the $K$ tables.
- Simulation 5:$\theta_{k} \eqd Skewed\text{-}Gaussian(\theta_{0}, 1, -10)$, where $\theta_{0} = -4, \ldots, 4$ is identical across the $K$ tables.
- Simulation 6: $\theta_{k} \eqd 0.5 \times Gaussian(- \theta_{0}, 1) + 0.5 \times Gaussian(\theta_{0}, 1)$.

Simulations 4a, 4b and 5 (Figures \ref{fig:Simulation_4a}, \ref{fig:Simulation_4b} and \ref{fig:Simulation_5}) demonstrate that the Mantel-Haenszel estimator, $\widehat{\theta}_{MH}$, is a reasonable estimator of the centre of the distribution of the true log odds ratios, $\theta_{k}$. It performs better in Simulation 4a and 4b, where the distribution of the $\theta_{k}$ is symmetric, as opposed to Simulation 5, where the distribution of the $\theta_{k}$ is skewed. Unsurprisingly, $\widehat{\theta}_{MH}$ does a better job when $K = 1000$ (Simulation 4b, Figure \ref{fig:Simulation_4b}) than when $K = 100$ (Simulation 4a, Figure \ref{fig:Simulation_4a}).

\begin{figure}[!htb]
\centering
\includegraphics[width=\textwidth]{../figures/Simulation_5.pdf}
\caption[Simulation 4a]{Results of simulation 4a ($K = 100$). Estimating a heterogeneous log odds ratio, $\theta_{k} = Gaussian(\theta_{0}, 1)$ (density shown by the dashed line), from a $2 \times 2 \times 100$ contingency table where the marginal probabilities, $\pi_{2+}$ and $\pi_{+2}$, are distributed as $Beta(0.3, 0.2)$ random variables. The results of three estimators of $\theta$ are shown: $\widehat{\theta}_{0.5}$ (histogram); the point estimate (solid line) and $95 \%$ confidence interval (dashed lines) of $\widehat{\theta}_{U}$ (coloured orange); and the point estimate (solid line) and $95 \%$ confidence interval (dashed lines) of $\widehat{\theta}_{MH}$ (coloured blue). The average sample size (sequencing-coverage) of each $2 \times 2$ table is $30$.}
\label{fig:Simulation_4a}
\end{figure}
